\documentclass[10pt,journal]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage{listings}
\usepackage{placeins}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  columns=fullflexible
}

\title{From Legal Provisions to Machine-Readable Controls:\\
A Transformation Framework for China AIGC Governance}

\author{Yuxuan Cheng%
\thanks{Research paper draft. This manuscript describes a research framework and does not claim legal compliance certification.}}

\IEEEtitleabstractindextext{%
\begin{abstract}
This paper presents a research framework for transforming selected Chinese AIGC legal provisions into traceable, reviewable, and machine-readable compliance controls. The project is motivated by a gap between natural-language AI governance obligations and the technical artifacts needed for repeatable compliance engineering. Rather than building another user-facing compliance assessment product, the framework studies how rules are produced from legal text: source clauses are selected, normative fragments are identified, legal semantics are decomposed, human interpretations are recorded, control objectives and evidence requirements are defined, and executable tests are attached with review status and traceability. The framework distinguishes fully automatable, partially automatable, and human-review-required obligations, and it supports pass, fail, review, and not-applicable outcomes. A compact synthetic actor and digital human use case is used only as proof of concept to test whether reviewed controls can be validated and executed. The contribution is methodological: machine-readable rules should execute human-reviewed legal interpretations, not replace legal interpretation.
\end{abstract}

\begin{IEEEkeywords}
AI governance, AIGC, compliance-as-code, legal informatics, machine-readable rules, traceability, synthetic actors, China AI regulation.
\end{IEEEkeywords}
}

\begin{document}
\maketitle
\IEEEdisplaynontitleabstractindextext
\IEEEpeerreviewmaketitle

\section{Introduction}

Chinese AIGC governance increasingly contains obligations that are operationally relevant to generative AI services, deep synthesis providers, synthetic actors, digital humans, AIGC labeling, model governance, content safety, complaint handling, and personal information protection. These obligations are written as legal and technical prose. Engineering teams, however, need controls, evidence fields, test logic, and review workflows. This creates a transformation problem: how can natural-language legal provisions become structured controls without pretending that software can interpret law by itself?

This paper reframes the China AIGC Compliance Evidence Generator as a legal-clause-to-control research framework. The project is distinct from a separate actor-compliance product demo that applies already-defined rules to concrete business facts. This repository studies the upstream method by which rules are derived from selected Chinese AIGC provisions.

The research questions are:

\begin{itemize}
  \item \textbf{RQ1:} Chinese AIGC legal and regulatory provisions can be decomposed into which semantic elements for machine-readable representation?
  \item \textbf{RQ2:} How can traceable mappings be established between source clauses, legal interpretations, control objectives, evidence requirements, and executable tests?
  \item \textbf{RQ3:} Which categories of legal obligations are fully automatable, partially automatable, or dependent on human legal review?
  \item \textbf{RQ4:} Can the proposed transformation method generate consistent and executable controls for a synthetic actor and digital human use case?
\end{itemize}

\section{Related Work}

Prior work on AI governance, model documentation, and risk management provides useful scaffolding for this project. Model Cards and Datasheets improve structured transparency, while the NIST AI Risk Management Framework provides vocabulary for governance and risk processes. OSCAL shows how security controls and assessment artifacts can be represented in machine-readable form. Compliance-as-code work demonstrates that operational requirements can sometimes be converted into executable checks. Rules-as-Code and computational-law projects study how public rules can be represented for digital administration. Legal ontology research formalizes legal concepts and relations at a higher semantic level.

This paper is narrower than those traditions. It does not propose a full ontology, a regulator-grade rule engine, or a replacement for legal interpretation. It focuses on the transformation step that often remains implicit: a natural-language Chinese AIGC clause is selected, interpreted by a human reviewer, decomposed into a structured norm, and then translated into an executable control. The contribution therefore sits between legal text and control execution:

\begin{lstlisting}
Natural-language legal clause
-> human-reviewed structured norm
-> executable control
\end{lstlisting}

Machine-readable AI compliance evidence projects often begin with a control catalog and then produce assessment artifacts. This project asks an upstream question: where did the control come from, and can that derivation remain reviewable?

\section{Research Gap}

Many AI compliance prototypes begin with controls and demonstrate whether a system passes or fails. That is useful for product demos, but it hides the origin of the controls. A reviewer may ask: Which legal clause produced this control? Which normative fragment was selected? Which legal interpretation was made? What evidence is required? Is the test fully automatable, partially automatable, or dependent on human review?

The gap is therefore not only evidence generation. It is traceable rule production. Without traceability, machine-readable controls can become detached from source law and may give a false impression of legal certainty.

\section{Regulatory Corpus and Clause Selection}

The project uses a small representative corpus rather than attempting full coverage of Chinese AI law. The selected sources include the Interim Measures for the Management of Generative AI Services, the Provisions on the Administration of Deep Synthesis Internet Information Services, the Measures for the Labeling of AI-Generated Synthetic Content, GB 45438--2025, and selected provisions from the Personal Information Protection Law.

These sources were selected because together they cover the main governance interfaces for China-facing AIGC actor and digital human services: public-facing generative AI services, deep synthesis services, explicit labels, metadata labels, filing triggers, complaint mechanisms, sensitive personal information, and face or voice editing. Clauses were included when they could be represented with a traceable source, actor, modality, action, object, trigger, exception, evidence requirement, automation level, and review status. Clauses were excluded when they were sector-specific, too broad for this prototype, unrelated to synthetic actor services, or dependent on extensive regulator practice not modeled here.

The corpus also preserves overlaps between general and special rules. For example, PIPL Article 29 is retained as a general sensitive-personal-information separate-consent rule, while Article 14 of the Deep Synthesis Provisions is retained as a special rule for face and voice editing functions. This avoids collapsing distinct legal sources into a single generic consent control.

\section{Legal-Clause-to-Control Transformation Method}

The transformation method follows a fixed chain:

\begin{lstlisting}
Formal Chinese legal provision
-> Normative sentence identification
-> Legal semantic decomposition
-> Human interpretation
-> Control objective
-> Evidence requirement
-> Machine-readable test
-> Review status and traceability
\end{lstlisting}

Each selected clause is decomposed using an annotation protocol. The required fields include formal Chinese source text, working English translation, official source URL, source document, article number, normative fragment, regulated actor, modality, action, object, trigger, exception, timing, required evidence, control objective, test logic, automation level, ambiguity level, review note, source effective date, author review status, legal expert review status, and interpretation version.

The protocol treats the Chinese source text as authoritative for the research artifact. English text is recorded only as a working translation. The schema no longer labels an English summary as original text. This distinction matters because a control should remain traceable to the formal legal source rather than to an informal summary.

Table~\ref{tab:semantic-fields} summarizes the main semantic elements.

\begin{table*}[!t]
\centering
\footnotesize
\caption{Semantic Elements Used for Clause Decomposition}
\label{tab:semantic-fields}
\begin{tabularx}{\textwidth}{@{}p{0.24\textwidth}p{0.64\textwidth}@{}}
\toprule
\textbf{Element} & \textbf{Purpose} \\
\midrule
Source and article & Preserve traceability to the legal or technical source. \\
Normative fragment & Identify the sentence or phrase that creates the obligation. \\
Regulated actor, action, object & Convert prose into a structured legal relation. \\
Trigger, exception, timing & Identify when the rule applies, when it does not apply, and when action is required. \\
Evidence requirement & Define the records needed before any machine-readable test can run. \\
Test logic & Attach an executable condition only after human interpretation. \\
Automation and ambiguity level & Prevent overclaiming by routing vague duties to review. \\
Review status & Distinguish author review from independent legal expert review. \\
\bottomrule
\end{tabularx}
\end{table*}

\section{Machine-Readable Schema}

The repository contains two schemas. The legal norm schema represents the human-reviewed interpretation of a clause. The control schema represents the executable control derived from that norm. Both require source traceability, evidence requirements, test logic, automation level, and review status. The source block includes Chinese title, English title, issuing authority, article, formal Chinese source text, working English translation, official source URL, promulgation date, effective date, and retrieval date.

\begin{lstlisting}
norm_id: CN-AIGC-LABEL-001
source:
  document_title_zh: 人工智能生成合成内容标识办法
  document_title_en: Measures for the Labeling of AI-Generated Synthetic Content
  article: 第四条 / 第九条
  source_text_zh: ...
  working_translation_en: ...
  source_url_specificity: document_page
  source_snapshot:
    verification_status: metadata_only
norm:
  regulated_actor: generative_ai_service_provider
  modality: must
  action: add_or_document_explicit_label
  object: user_facing_generated_content
interpretation:
  automation_level: fully_automatable
evidence:
  required_fields:
    - visible_label_present
test:
  expression: visible_label_present == true
review:
  author_review_status: reviewed
  legal_expert_review_status: not_reviewed
\end{lstlisting}

The schema is intentionally modest. It does not encode all legal meaning. It checks whether a proposed rule contains the minimum information needed for traceability and execution. Source metadata validation checks that required source fields are present and internally consistent; it does not automatically prove that the stored text remains identical to the current official webpage. The prototype therefore records `metadata_only` source-snapshot status unless independent snapshot or manual verification is recorded.

\section{Human Review and Traceability}

The framework explicitly rejects the idea that machine-readable means machine-interpreted. The machine executes a reviewed representation. Human review remains necessary for selecting clauses, identifying normative fragments, interpreting ambiguous obligations, determining applicability, and deciding whether evidence is legally sufficient.

The initial interpretation in this prototype is performed by the author as a research exercise. The repository therefore separates \texttt{author\_review\_status} from \texttt{legal\_expert\_review\_status}. A status of ``reviewed'' currently means internally reviewed by the researcher unless a qualified legal expert review is separately recorded. Expert review is especially important for public-opinion or social-mobilization capacity, sensitive personal information classification, label prominence, and obligations that use open-textured terms such as effective measures.

Rules are classified into three automation levels. Metadata field presence may be fully automatable once applicability is established. Human-confirmed applicability is represented as a structured record that identifies the confirmed value, reviewer role, and review time, and is checked against business evidence by the evaluator. Label prominence, confusion risk, and content-safety adequacy are routed through a declared human-review key because legal adequacy cannot be reduced to a single Boolean field.

\subsection{Direct Obligations and Derived Organizational Controls}

The framework distinguishes a direct legal obligation from a supporting organizational assurance control. The first maps the legal duty itself into a machine-readable test. The second records an evidence practice that an engineering organization may choose in order to support assurance. For Deep Synthesis Article 14, the direct control checks that the provider presents a user prompt concerning notice and separate consent. A separate derived control checks whether edited-person consent evidence and scope review have been retained. The latter is not presented as a verbatim statutory requirement that a provider must retain a particular record in every case. Each control records `control_derivation_type` and `control_component` to make this relationship visible.

\subsection{Lifecycle Scope}

The Article 9 exception in the Labeling Measures is scoped to provider delivery to the requesting user. A pass only means that the prototype found the specified delivery-exception evidence. It does not decide later publication, user declaration, dissemination-platform identification, or platform notice obligations. The prototype deliberately does not automate the complete dissemination chain.

\section{Case Study: Synthetic Actor and Digital Human Rules}

The case study includes controls for visible AIGC label existence, visible-label prominence, metadata labels, GB metadata fields, PIPL separate consent, biometric-like classification, deep synthesis face and voice editing consent, model filing applicability, content-safety principles, complaint mechanisms, and deep synthesis confusion-risk labeling.

Table~\ref{tab:control-types} shows representative controls.

\begin{table*}[!t]
\centering
\footnotesize
\caption{Representative Clause-to-Control Mappings}
\label{tab:control-types}
\begin{tabularx}{\textwidth}{@{}p{0.22\textwidth}p{0.22\textwidth}p{0.22\textwidth}p{0.18\textwidth}@{}}
\toprule
\textbf{Norm} & \textbf{Control} & \textbf{Evidence} & \textbf{Automation} \\
\midrule
Visible AIGC label & Label existence and Article 9 exception evidence & labeling check & partially automatable \\
Label prominence & Prominent placement and perceptibility & human label review & human review required \\
Metadata label & Underlying metadata fields exist & label metadata & fully automatable \\
Separate consent & Sensitive actor-like data has separate consent & consent record & partially automatable \\
Face and voice editing & Direct user prompt / derived consent assurance & prompt and assurance records & partially automatable \\
Model filing & Filing applicability and registered metadata are checked & model filing record & partially automatable \\
Content safety & Article 4 principle and prohibited-content review & safety review record & human review required \\
\bottomrule
\end{tabularx}
\end{table*}

\section{Prototype Validation}

Validation is intentionally small and explainable. The repository includes a compact set of validation cases covering applicable evidence, missing evidence, clear failures, non-applicability, human-review-required obligations, explicit-label exceptions, filing non-applicability, unreviewed filing applicability, simultaneous PIPL and deep synthesis consent duties, and metadata aggregate flags that mask missing underlying fields. The goal is not to report a pass/fail distribution. The goal is to test whether clauses can be consistently structured, controls can be traced to source provisions, machines can execute clear conditions, ambiguous duties are routed to review, and schemas detect incomplete rule definitions.

The validation method has five layers. First, schema completeness checks whether complete norm artifacts and controls contain source, review, evidence, and test fields. Second, source traceability checks whether each control can be linked back to a source identifier, official URL, article, effective date, and URL-specificity classification. Third, cross-file consistency checks whether source index, norm artifacts, clause decomposition, traceability matrix, and mapping agree on norm identifiers, article numbers, titles, dates, automation level, and interpretation version. Fourth, executable coverage checks whether expression variables are declared in evidence, applicability, exception, confirmation, or review structures. Fifth, review-routing accuracy checks whether human-confirmed applicability and keyed human review prevent premature final pass results.

Error analysis is part of the method. The validation set includes cases for incorrect source mapping, missing source text, undeclared expression variables, incomplete explicit-label exception evidence, unreviewed filing applicability, and metadata controls where \texttt{metadata\_label\_present} is true but required underlying fields are missing.

The previous 1,000-request synthetic pipeline remains as supplementary proof of concept. It demonstrates that structured controls can be read and exercised against synthetic operational records, but it is no longer the central result of the paper.

\section{Limitations}

The corpus is small and illustrative. Clause summaries are not official translations. The schema is not a legal ontology. The evaluator is deliberately limited and should not be used as a production compliance engine. Automation classifications are research judgments that require legal review before operational use. The project uses no real personal data, biometric data, face data, voice data, contracts, company records, or production media. It does not certify compliance with any law or standard.

\section{Conclusion}

This paper presents a traceable method for transforming selected Chinese AIGC legal provisions into machine-readable controls. The contribution is not a product that decides compliance. It is a research framework for preserving the path from legal text to reviewed interpretation, evidence requirement, executable test, automation level, and review status. Future work should expand legal review, compare alternative interpretations, and integrate stronger provenance for rule-authoring decisions.

\end{document}
